Finding shortest and closest vectors in a lattice of Voronoi's first kind

### 2014-05-26 talk/presentation.tex
minor

\frame{
\frametitle{The shortest vector problem}
\begin{itemize}
\item Computing a short vector is called the \term{shortest vector problem}.
\item Applications in cryptography and number theory.
\item NP-hard for arbitrary lattices.
\item Easier for specific lattices.
\item For example, short vectors are easy to find in the \term{root lattices} $\ints^n$, $A_n$, and $D_n$.
\item We will show that the problem is relatively easy to solve for lattices of \term{Voronoi's first kind}.
\end{itemize}
}

@@

\frame{ 
\frametitle{Relevant vectors}
\begin{itemize}
\item The relevant vectors are those $v \in \Lambda \backslash \{0\}$ such that  
\[
\dotprod{v}{x} \leq \dotprod{x}{x} \qquad \text{for all $x \in \Lambda$.}
\]
\item The Voronoi cell can be defined using the relevant vectors, 
\[
\vor(\Lambda) = \{ x \in\reals^m \mid \|x\| \leq \|x - v\|, v \in\relevant(\Lambda) \}.
\]
\item Short vectors are relevant vectors.
\end{itemize}
}

@@

\frame{
\frametitle{The closest lattice point problem}
Applications to:
\begin{itemize}
\item coding and quantisation,
\item multi-antenna communications (MIMO),
\item unwrapping of phase data for electronic distance measurement in GPS and surveying,
\item single frequency estimation,
\item polynomial phase estimation,
\item circular statistics.
\end{itemize}
}

\frame{
\frametitle{The closest lattice point problem}
\begin{itemize}
\item NP-hard for arbitrary lattices.
\item Easier for specific lattices.
\item For example, fast algorithms exist for the \term{root lattices} $\ints^n$, $A_n$, and $D_n$.
\item We will describe a fast algorithm to compute a closest point in lattices of \term{Voronoi's first kind}.
\end{itemize}
}

@@

\frame{
\frametitle{Lattices of Voronoi's first kind}
\begin{theorem}
Let $\Lambda$ be a $n$-dimensional lattice of Voronoi's first kind with obtuse superbase $b_1, \dots, b_{n+1}$.  The relevant vectors in $\Lambda$ are of the form 
\[
\sum_{i \in I} b_i
\] 
where $I$ is a strict subset of $\{1, 2, \dots, n+1\}$ and $I$ is not empty.
\end{theorem}

@@

\frame{
\frametitle{Graphs, cuts, and minimum cuts}
If the edge weights $w_{ij}$ are all nonnegative, a minimum cut can be computed:
\begin{itemize}
\item deterministically in $O(n^3)$ operations using the algorithm of Stoer and Wagner,
\item with high probability in $O(n^2 \log(n)^3)$ operations using the randomised algorithm of Karger and Stien.
\end{itemize}	
}

\frame{
\begin{theorem}
Let $\Lambda$ be a $n$-dimensional lattice of Voronoi's first kind with obtuse superbase 
\[
b_1, \dots, b_{n+1}.
\]  
Let $G$ be a graph with $n+1$ vertices $v_{1}, \dots, v_{n+1}$ and edge weights 
\[
w_{ij} = -q_{ij} = -b_i \cdot b_j \geq 0 \qquad i \neq j.
\]  
Let $(C, \bar{C})$ be a minimum cut in $G$.  A short vector in $\Lambda$ is 
\[
\sum_{i \in I} b_i \;\;\; \text{where} \;\;\; I = \{ i \mid v_i \in C\}.
\]  
The squared length of the short vector is given by the weight of the minimum cut. %$W(C,\bar{C})$.
\end{theorem}
}

@@

\frame{
\frametitle{A series of relevant vectors}
For a lattice of Voronoi's first kind:
\begin{itemize}
\item $x_0$ can be chosen to ensure that a closest lattice point in found after at most $n$ iterations.
\item Minimisation over the set of relevant vectors can be performed by computing a minimum cut in a flow network.
\item Using known algorithms a minimum cut can be found in $O(n^3)$ operations.
\item In total $O(n^4)$ operations are required to compute a closest lattice point.
\end{itemize}
}

@@

\begin{theorem}
Let $\Lambda$ be a $n$-dimensional lattice of Voronoi's first kind with obtuse superbase $b_1, \dots, b_{n+1}$.  Let $z_1,\dots,z_{n+1} \in \reals$ minimise
\[
\| y - \sum_{i=1}^{n+1} b_i z_i \| 
\]
and put
\[
x_0 = \sum_{i=1}^{n+1} b_i \floor{z_i}.
\]
The iterative procedure, initialized at $x_0$, converges to a closest lattice point in at most $n$ iterations.
\end{theorem}
}

@@

\begin{itemize}
\item Can good codes or quantisers be constructed from lattices of Voronoi's first kind?
\item Do applications such as global position, phase unwrapping, circular statistics, etc., involve lattices of Voronoi's first kind?
\item Are there subfamilies of Voronoi's first kind that admit even faster algorithms?
\item Are there other families of lattices for which similar techniques lead to polynomial time algorithms?
\end{itemize}

@@

>>>>

\frame{
\frametitle{The shortest vector problem}
\begin{itemize}
\item Computing a short vector is called the \term{shortest vector problem}.
\item Applications in cryptography and number theory.
\item NP-hard for arbitrary lattices.
\item Easier for specific lattices.
\item Short vectors are easy to find in the \term{root lattices} $\ints^n$, $A_n$, and $D_n$.
\item We will show that the problem is relatively easy to solve for lattices of \term{Voronoi's first kind}.
\end{itemize}
}

@@

\frame{ 
\frametitle{Relevant vectors}
\begin{itemize}
\item Denote the set of relevant vectors by $\relevant(\Lambda)$.
%\item The relevant vectors are those $v \in \Lambda \backslash \{0\}$ such that  
%\[
%\dotprod{v}{x} \leq \dotprod{x}{x} \qquad \text{for all $x \in \Lambda$.}
%\]
\item The Voronoi cell can be defined using the relevant vectors, 
\[
\vor(\Lambda) = \{ x \in\reals^m \mid \|x\| \leq \|x - v\|, v \in\relevant(\Lambda) \}.
\]
\item Short vectors are relevant vectors.
\end{itemize}
}

@@

\frame{
\frametitle{The closest lattice point problem}
Applications to:
\begin{itemize}
\item coding and quantisation,
\item multi-antenna communication (MIMO),
\item unwrapping of phase data for electronic distance measurement in GPS and surveying,
\item single frequency estimation,
\item polynomial phase estimation,
\item circular statistics.
\end{itemize}
}

\frame{
\frametitle{The closest lattice point problem}
\begin{itemize}
\item NP-hard for arbitrary lattices.
\item Easier for specific lattices.
\item Fast algorithms exist for the \term{root lattices} $\ints^n$, $A_n$, and $D_n$.
\item We will describe a fast algorithm to compute a closest point in lattices of \term{Voronoi's first kind}.
\end{itemize}
}

@@

\frame{
\frametitle{Lattices of Voronoi's first kind}
\begin{theorem}[Conway and Sloane (1992)]
Let $\Lambda$ be a $n$-dimensional lattice of Voronoi's first kind with obtuse superbase $b_1, \dots, b_{n+1}$.  The relevant vectors in $\Lambda$ are of the form 
\[
\sum_{i \in I} b_i
\] 
where $I$ is a strict subset of $\{1, 2, \dots, n+1\}$ and $I$ is not empty.
\end{theorem}

@@

\frame{
\frametitle{Graphs, cuts, and minimum cuts}
If the edge weights $w_{ij}$ are all nonnegative, a minimum cut can be computed:
\begin{itemize}
\item deterministically in $O(n^3)$ operations using the algorithm of Stoer and Wagner (1997),
\item with high probability in $O(n^2 \log(n)^3)$ operations using the randomised algorithm of Karger and Stien (1996).
\end{itemize}	
}

\frame{
\begin{theorem}[McKilliam and Grant (2012)]
Let $\Lambda$ be a $n$-dimensional lattice of Voronoi's first kind with obtuse superbase 
\[
b_1, \dots, b_{n+1}.
\]  
Let $G$ be a graph with $n+1$ vertices $v_{1}, \dots, v_{n+1}$ and edge weights 
\[
w_{ij} = -q_{ij} = -b_i \cdot b_j \geq 0 \qquad i \neq j.
\]  
Let $(C, \bar{C})$ be a minimum cut in $G$.  A short vector in $\Lambda$ is 
\[
\sum_{i \in I} b_i \;\;\; \text{where} \;\;\; I = \{ i \mid v_i \in C\}.
\]  
The squared length of the short vector is given by the weight of the minimum cut. %$W(C,\bar{C})$.
\end{theorem}
}

@@

\frame{
\frametitle{A series of relevant vectors}
For a lattice of Voronoi's first kind:
\begin{itemize}
\item $x_0$ can be chosen to ensure that a closest lattice point is found after at most $n$ iterations.
\item Minimisation over the set of relevant vectors can be performed by computing a minimum cut in a flow network.
\item Using known algorithms a minimum cut can be found in $O(n^3)$ operations.
\item In total $O(n^4)$ operations are required to compute a closest lattice point.
\end{itemize}
}

@@

\begin{theorem}[McKilliam, Grant, Clarkson (2014)]
Let $\Lambda$ be a $n$-dimensional lattice of Voronoi's first kind with obtuse superbase $b_1, \dots, b_{n+1}$.  Let $z_1,\dots,z_{n+1} \in \reals$ minimise
\[
\| y - \sum_{i=1}^{n+1} b_i z_i \| 
\]
and put
\[
x_0 = \sum_{i=1}^{n+1} b_i \floor{z_i}.
\]
The iterative procedure, initialized at $x_0$, converges to a closest lattice point in at most $n$ iterations.
\end{theorem}
}

@@

\begin{itemize}
\item Can good codes or quantisers be constructed from lattices of Voronoi's first kind?
\item Do applications such as global positioning, phase unwrapping, etc., involve lattices of Voronoi's first kind?
\item Are there subfamilies of Voronoi's first kind that admit even faster algorithms?
\item Are there other families of lattices for which similar techniques lead to polynomial time algorithms?
\end{itemize}

@@

% \frame{

% \frametitle{An extra problem}

% Choose real numbers $\lambda_1,\dots,\lambda_n$ to minimise
% \[
% \sum_{i=1}^n \frac{P^2}{\lambda_i^2}
% \]
% where $P  = \operatorname{lcm}(\lambda_1,\dots,\lambda_n)$ subject to the constraints
% \[
% P > P_{\text{min}}
% \]
% and 
% \[
% \lambda_{\text{min}} < \lambda_i < \lambda_{\text{max}} \qquad i = 1,\dots,n.
% \]

% }
